Inference for Dynamic and Latent Variable Models via Plug-and-Play Automatically Differentiable Particle Filtering

### 2023-11-03 paper/ms.tex
more on paper/ms

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{Please provide details of author contributions here.}
\authordeclaration{Please declare any competing interests here.}
%\equalauthors{\textsuperscript{1}A.O.(Author One) contributed equally to this work with A.T. (Author Two) (remove if not applicable).}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: ionides@umich.edu}

@@

\begin{abstract}
  Automatic differentiation has driven recent advances in machine learing, including deep neural networks and Hamiltonian Markov Chain Monte Carlo methods.
  This progress has required simultaneous advances in algorithms, software and hardware.  
  Partially observed nonlinear stochastic dynamic systems have proved resistant to these techniques; despite various attempts, widely applicable methods have not yet emerged.
  We present a new approach which is applicable to a general class of models, possesses a theoretical foundation, and is demonstrated to beat current state-of-the-art methods on a challenging scientific benchmark problem.
  Our algorithm is compatible with parallel computation on a graphical processing unit, and enjoys the plug-and-play property that its software implementation requires only a simulator for the scientific dynamic model as input (in addition to data, and a measurement model).

@@

\firstpage{7}
% Use \firstpage to indicate which paragraph and line will start the second page and subsequent formatting. In this example, there are a total of 11 paragraphs on the first page, counting the first level heading as a paragraph. The value {12} represents the number of the paragraph starting the second page. If a paragraph runs over onto the second page, include a bracket with the paragraph line number starting the second page, followed by the paragraph number in curly brackets, e.g. "\firstpage[4]{11}".

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{M}any approaches to inference in highly nonlinear stochastic dynamical systems assume access to the probability density of next states given the current state.

This is a problem in some critical applications, such as disease modeling, where the models are complex enough that obtaining the density is intractable. However, the particle filter, a popular method for solving the filtering problem in partially-observed dynamical systems, does not require evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm in a feature known as the plug-and-play property.

@@

We propose a hybrid algorithm that u

Unlike IF2, we explicitly characterize our method's rate of convergence

@@

\subsection*{Tables}
Tables should be included in the main manuscript file and should not be uploaded separately.

\subsection*{Single column equations}

Authors may use 1- or 2-column equations in their article, according to their preference.

To allow an equation to span both columns, use the \verb|\begin{figure*}...\end{figure*}| environment mentioned above for figures.

Note that the use of the \verb|widetext| environment for equations is not recommended, and should not be used.

\begin{figure*}[bt!]
\begin{align*}
(x+y)^3&=(x+y)(x+y)^2\\
       &=(x+y)(x^2+2xy+y^2) \numberthis \label{eqn:example} \\
       &=x^3+3x^2y+3xy^3+x^3.
\end{align*}
\end{figure*}

>>>>

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{K.T. and E.L.I. planned the study; K.T. developed the numerical results; all authors wrote the manuscript.}
\authordeclaration{The authors declare no competing interests.}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: ionides@umich.edu}

@@

\begin{abstract}
  Automatic differentiation has driven recent advances in machine learing, including deep neural networks and Hamiltonian Markov Chain Monte Carlo methods.
  This progress has required simultaneous advances in algorithms, software and hardware.  
  Partially observed nonlinear stochastic dynamic systems have proved resistant to the benefits of automatic differentiation techniques; despite various attempts, widely applicable methods have not yet emerged.
  We present a new approach which is applicable to a general class of models, possesses a theoretical foundation, and is demonstrated to beat current state-of-the-art methods on a challenging scientific benchmark problem.
  Our algorithm is compatible with parallel computation on a graphical processing unit, and enjoys the plug-and-play property that its software implementation requires only a simulator for the scientific dynamic model as input (in addition to data, and a measurement model).

@@

\firstpage{3}
% Use \firstpage to indicate which paragraph and line will start the second page and subsequent formatting. In this example, there are a total of 11 paragraphs on the first page, counting the first level heading as a paragraph. The value {12} represents the number of the paragraph starting the second page. If a paragraph runs over onto the second page, include a bracket with the paragraph line number starting the second page, followed by the paragraph number in curly brackets, e.g. "\firstpage[4]{11}".

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{T}he particle filter is a widely used Monte Carlo algorithm which provides an unbiased estimate of the likelihood function for partially observed Markov process (POMP) models \ed{REFS}.
Maximization of the likelihood permits computation of the maximum likelihood parameter estimate, as well as profile likelihood confidence intervals, and model selection by likelihood ratio tests or Akaike's information criterion.
The particle filter involves a random resampling step in which particles inconsistent with the data are likely to be pruned out and the most consistent particles are replicated.
Applying automatic differentiation (AD) to supply the optimization with gradient information is complicated by the disctete nature of this resampling step.
Simulations of discrete random variables are necessarily discontinuous, specifically piecewise constant, as a function of the unknown parameters for a fixed value of the seed for the random number generator.
As a result of this, the Monte Carlo expectation of the derivative is not equal to the derivative of the Monte Carlo expectation.
Various strategies have been proposed to address this phenomenon \ed{REFS}, but none of these algorithms have found widespread practical applicability.

A significant practical strength of the particle filter is its applicability to arbitrarily nonlinear stochastic systems, requiring only a simulator for the dynamic model.
This is called the plug-and-play property \ed{refs}, and this property is critical in some applications, such as disease modeling, where the models are complex enough that obtaining the density is intractable.
Some extensions of the particle filter to parameter estimation require availability of transition probabilities for the dynamic model \ed{refs} which violates the plug-and-play property.
We have developed an AD algorithm for the particle filter which preserve the plug-and-play property while providing useful derivative estimates which demonstrably improve our ability to search for the maximum of the likelihood.

Our key insights are as follows:
\begin{enumerate}
\item We consider a novel one-parameter family of plug-and-play AD gradient estimators allowing flexibility in the tradeoff between Monte Carlo bias and variance. For our target problem, the results are insensitive to the choice of this parameter.

\item ADPF stochastic gradient optimization has complementary strengths to iterated filtering algorithms that are widely used for plug-and-play likelihood maximization for POMP models. Iterated filtering algorithms provide a computationally efficient way to approach the maximum, but they are poor at identifying the precise location of the maximum.
Combining both these algorithms substantially out-performs either alone.

\item The IFAD algorithm can be implemented on a GPU to achieve massive parallelization. 
\end{enumerate}
Together, these innovations lead to methodology which advances capability for statistical analysis of scientific models arising in biological sciences and elsewhere \ed{refs}.
Our benchmark task is a model previously used to evaluate iterated filtering algorithms \ed{refs}.

\subsection*{Assorted draft stuff}

[naive ADPF] intruduces bias into the Monte Carlo estimate of the derivative.
This bias has previously been reported to be non-negligible [REF].
An approach to avoiding this bias is the REINFORCE estimator, also known as the stop-gradient trick \cite{scibior21}.
The Monte Carlo variance of this estimator has been found to be problematic \cite{scibior21}

1. A new derivation of the stop-gradient estimator. We show that this estimator can be obtained as the fixed seed derivative of an extension of the basic particle filter which we call the off policy particle filter.

2. Combining the reparameterization trick with the stop-gradient estimator to obtain a plug-and-play gradient estimator.

3. Apply this gradient estimator within a stochastic gradient descent algorithm using a normalized gradient.

4. Multiple searches from diverse starting points to reduce, quantify and control maximization error due to multimodality and Monte Carlo variability.

%\dropcap{M}
Many approaches to inference in highly nonlinear stochastic dynamical systems assume access to the probability density of next states given the current state.

@@

%We propose a hybrid algorithm that u

%Unlike IF2, we explicitly characterize our method's rate of convergence

@@

%% \subsection*{Tables}
%% Tables should be included in the main manuscript file and should not be uploaded separately.

%% \subsection*{Single column equations}

%% Authors may use 1- or 2-column equations in their article, according to their preference.

%% To allow an equation to span both columns, use the \verb|\begin{figure*}...\end{figure*}| environment mentioned above for figures.

%% Note that the use of the \verb|widetext| environment for equations is not recommended, and should not be used.

%% \begin{figure*}[bt!]
%% \begin{align*}
%% (x+y)^3&=(x+y)(x+y)^2\\
%%        &=(x+y)(x^2+2xy+y^2) \numberthis \label{eqn:example} \\
%%        &=x^3+3x^2y+3xy^3+x^3.
%% \end{align*}
%% \end{figure*}
